Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-01 doc/invt.tex
invt.tex: update the theory

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
The method reproduces the inverse-time formula
for the WL case.
%
It also predicts, for metadynamics,
a more algebraically complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian wave packet.
%
Further, by comparing
different updating schemes,
we show that
the single-bin updating scheme
used by the WL algorithm
can be generalized to class of multiple-bin updating schemes
that are partially optimal for infinitely long simulations,
and the inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

As shown above, with fixed endpoint value, $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
  \label{eq:error_asym_Lagrangian}
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root of the left-hand side, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
  %\label{eq:q_opt}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.

@@

@@

It is natural to introduce a homogeneous updating scheme
for a periodic variable\cite{dama2014}.
%
Consider the following matrix, $\mathbf w$,
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b \le n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $l > b$.
%
Particularly,
we shall only consider symmetric kernels
that satisfy
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
If $b = n/2$, then $\mu_{-b}$ is dropped
from Eq. \eqref{eq:normalization}.
%
These numbers constitute a \emph{updating kernel}.

@@

We shall first show,
directly from Eq. \eqref{eq:error_asym1},
that the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence same $\Gamma_k$'s,
but with different eigenvalues, $\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of nonnegative numbers, $c_k$'s,
%
\begin{align}
&
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr)
\right)
%
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k \, \dot u_k \, \bigl( q(t) \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 2}^n \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k \, q(T) }
    \right]
\right)^2.
\label{eq:CSineq}
\end{align}
%
This inequality follows from the non-negativity of
the following quadratic polynomial of $Y$,
$$
\int_0^T
  dt \sum_{k = 2}^n \Gamma_k \,
    \left( \dot u_k \, Y - c_k \right)^2
  \equiv
  A \, Y^2 + B \, Y + C
%\ge 0,
$$
which implies a non-positive discriminant,
$B^2 - 4 \, A \, C$.
%
Note that the last expression of \eqref{eq:CSineq}
depends on the curve, $q(t)$ ($0 < t < T$),
only through the endpoint value, $q(T)$,
which is fixed in the variation process.
%
Thus, the inequality sets a lower bound
of the error $\Err$
given in Eq. \eqref{eq:error_asym1}.
%
The equality is achieved
if $\dot u_k\left( q(t) \right) = c_k$
for all $k = 2, \dots, n$ at any $t$,
up to a multiplicative factor.
%
Solving this set of equations yields
%
$$
\alpha(t) = \frac{              1             }
                 { \lambda_k \, (t + t_{k0} ) },
\quad
\mathrm{with\;\;}
t_{k0} = \frac{             T            }
              { e^{ \lambda_k q(T) } - 1 }.
$$
%
Such a solution is possible only if
%
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{We add some steps for the above solution.
  Integrating $\dot u_k = c_k$ yields
  \begin{equation}
  u_k(t) = c_k \left(t + t_{k0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_k(T) = 1$ and $u_k(0) = e^{-\lambda_k \, q(T)}$,
  we have
  $$
  c_k = \frac{ 1 }{ T + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { T + t_{k0} }
  =
  e^{ -\lambda_k \, q(T) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_k \, q(T) + \lambda_k \, q(t)
  = \log c_k + \log\left( t + t_{k0} \right).
  $$
  Differentiating this with respect to $t$ yields
  $$
  \lambda_k \, \alpha(t)
  =
  \frac {     1      }
        { t + t_{k0} }.
  $$
}

Since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_k$,
we can freely set $\lambda_1$ to $\lambda_2$
without changing the nature of the updating scheme.
%
Now an updating matrix $\mathbf w$
with identical eigenvalues
is essentially a multiple of the identity matrix.
%
Thus, if no eigenvalue $\lambda_k$ is zero,
then in terms of asymptotic convergence,
the single-bin scheme adopted in the WL algorithm
is most efficient.
%
In the single-bin case,
$c_k = 1/(T + t_0)$,
and the error is
\begin{align}
  E
  =
  \frac{       T     }
       { (T + t_0)^2 }
  \sum_{ k = 2 }^n
    \Gamma_k
  .
\label{eq:error_singlebin}
\end{align}

Note that the above result
holds only under the long time limit,
and it is not necessarily true
for finite length simulations.
%
For example, consider a system that is initially
prepared under a fixed $\alpha_0$
at the beginning of the schedule $\alpha(t)$.
%
Then for shorter times, $T \approx 0$,
the error, given by Eq. \eqref{eq:error_eql},
is clearly smaller if some eigenvalues
are less than $1.0$ (the single-bin case).
%
%In practice, real simulations
%may have long achieved the desired precision
%before reaching the asymptotic regime,
%especially if the updating matrix
%has some $\lambda_k$'s close to $0$.

\subsubsection{\label{sec:optscheme}
Optimal schemes}

We can also generalize
the single-bin scheme to a class of
optimal updating schemes.
%
For example,
if we know in advance that the intrinsic
PMF is sufficiently smooth such that
the error modes can be limited to $k \le K$.
%
Then we can manually construct
an updating scheme with
$$
\lambda_1 = \cdots = \lambda_K = 1,
\mathrm{\; and \;}
\lambda_{K+1} = \cdots = \lambda_n = 0.
$$
For homogeneous updating schemes,
the updating kernel can be inversely
found from Eq. \eqref{eq:mu_from_lambda_pbc}
or \eqref{eq:mu_from_lambda_refl}.
%
Then,
Eq. \eqref{eq:error_singlebin} is modified to
$$
  E
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 2 }^K
    \Gamma_k.
$$

The optimal schedule of these schemes
remains to be the inverse-time schedule,
given by Eq. \eqref{eq:alpha_invt1}.

@@

To make better comparison with
the error of the single-bin case,
we shall rewrite the error as
a sum over the eigenmodes.
We define
\begin{align}
  \Theta_k
  =
  \frac{ 1 } { T }
  \sum_{ t = 1 } ^ T
    \theta_k( t )
  ,
\notag
\end{align}
%
with $\theta_k(t_1)$ defined by Eq. \eqref{eq:eta_def}
and $\phi_{k i}$ by Eq. \eqref{eq:eig_orthonormal_cols}.
%\begin{align}
%\theta_k( t_1 )
%=
%\sum_{ i = 1 }^ n
%\phi_{k i} \left[
%             \frac{ h_i(t_1) }
%                  { p_i      }
%             - 1
%           \right],
%\end{align}
%and
%\begin{align}
%\sum_{ k = 1 }^n
%  \phi_{k i} \, \phi_{k j}
%  = p_i \, \delta_{i j}.
%\end{align}
%
Then,
%
\begin{align}
  E
  =
  \sum_{ k = 1 }^n
    \left\langle
      \Theta_k^2
    \right\rangle.
\notag
%\label{eq:error_sbin_equil}
\end{align}
%
\note{Because
%
\begin{align*}
\sum_{ k = 1 }^n
  \Theta_k^2
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ k = 1 }^n
    \theta_k( t ) \, \theta_k( t' )
\\
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ i, j = 1 }^n
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
    \left[
      \frac{ h_j(t') }
           { p_j     }
      - 1
    \right]
    \sum_{ k = 1 }^n
      \phi_{k \, i} \, \phi_{k \, j}
\\
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ i, j = 1 }^n
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
    \left[
      \frac{ h_j(t') }
           { p_j     }
      - 1
    \right]
    p_i \, \delta_{i, j}
\\
&=
\sum_{ i = 1 }^n
\frac{ 1 } { T }
\sum_{ t = 1 }^T
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
\frac{ 1 } { T }
\sum_{ t' = 1 }^T
    \left[
      \frac{ h_i(t') }
           { p_i     }
      - 1
    \right]
\\
&=
\sum_{ i = 1 }^n
  p_i \,
    \left(
      \frac{ \bar h_i }
           { p_i      }
      - 1
    \right)^2
.
\end{align*}
}
The sum can start from $k = 2$, for
by Eq. \eqref{eq:eigenmode1}, we have
$$
\theta_1(t)
=
\sum_{ i = 1 }^n
  \phi_{1i} \, \frac{ h_i(t) - p_i } { p_i }
=
\sum_{ i = 1 }^n h_i(t)
-
\sum_{ i = 1 }^n p_i = 0.
$$

@@

In conclusion,
we have proposed a method of computing
the optimal schedule of the updating magnitude
for a general adaptive FDS simulation.
%
Adaptive FDS methods differ by the updating schemes,
which specify how the bias potential is modified
in each updating step.
%
In the single-bin or WL case,
the update is limited to the current bin;
while in the metadynamics case,
the bias potential of a few neighboring bins to the current one
is updated according to a Gaussian wave packet.

>>>>

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
The method reproduces the inverse-time formula
for the WL case.
%
It also predicts, for metadynamics,
a more algebraically complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian wave packet.
%
Further, we show that
the single-bin updating scheme
used by the WL algorithm
can be generalized to class of multiple-bin updating schemes
that are partially optimal for infinitely long simulations,
and the inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.

The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

As shown above, with fixed endpoint value, $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
  \label{eq:error_asym_Lagrangian}
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root of the left-hand side, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
  %\label{eq:q_opt}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error $E_A$,
the parameter $q(T)$ can be further optimized
such that total error, $E = E_R + E_A$, is minimal.

@@

\subsubsection{\label{sec:program}
Program}

We can now give a complete program of computing
the optimal schedule for a given updating scheme.

\begin{enumerate}

\item
Given an updating matrix, $\mathbf w$,
compute the eigenvalues, $\lambda_k$'s.

\item
Run a preliminary adaptive FDS simulation
using the updating scheme with a constant
updating magnitude, $\alpha_0$.
Use Eq. \eqref{eq:varY} to compute $\Gamma_k$'s.

\item
Select an initial value of $q(T)$,
e.g., $\ln(1 + \alpha_0 \, T)$.

\item
Compute the optimal schedule from
numerically integrating Eq. \eqref{eq:q_opt}.

\item
Compute the error $E$ from
Eqs. \eqref{eq:error_split},
  \eqref{eq:error_asym1} and
  \eqref{eq:error_res1}.

\item
Modify the value of $q(T)$ and repeat the above two steps,
until the minimal error $E$ is found.

\end{enumerate}

The dependence of variables are shown in Fig. \ref{fig:vardep}.

\begin{figure}
\begin{tikzpicture}
  \node[draw, text width={width("Updating ")}]
    (w) at (0, 3) {Updating \\matrix, $w$};

  \node[draw, text width={width("System  11")}]
    (sys) at (0, 1) {System \& \\sampling \\process};

  \node[draw, text width={width("Eigenvalues, ")}]
    (lambda) at (2.5, 3) {Eigenvalues, \\$\lambda_2, \dots, \lambda_n$};

  \node[draw, text width={width("Integrals, ")}]
    (gamma) at (2.5, 1) {Integrals, \\$\Gamma_2, \dots, \Gamma_n$};

  \node[draw, text width={width("Simulation ")}]
    (T) at (2.5, -0.5) {Simulation \\time, $T$};

  \node[draw, text width={width("Schedule ")}]
    (alpha) at (5, 1) {Optimal schedule, \\ $\alpha(t)$};

  \node[draw, text width={width("Error ")}]
    (err) at (7, 1) {Error, \\ $\Err$};

  \draw[->] (w) edge (lambda)
            (w) edge (gamma)
            (sys) edge (gamma)
            (lambda) edge (alpha)
            (gamma) edge (alpha)
            (T) edge (alpha)
            (alpha) edge (err);

\end{tikzpicture}
\caption{
  \label{fig:vardep}
  Dependence of variables in the method of computing
  the optimal schedule.
}
\end{figure}

@@

It is natural to introduce a homogeneous updating scheme
for a periodic variable\cite{dama2014}.
%
Consider the following matrix, $\mathbf w$,
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b \le n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $l > b$.
%
Particularly,
we shall only consider symmetric kernels
that satisfy
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
If $b = n/2$, then $\mu_{-b}$ is dropped
from Eq. \eqref{eq:mu_normalization}.
%
These numbers constitutes a \emph{updating kernel}.

@@

We shall first show,
directly from Eq. \eqref{eq:error_asym1},
that the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence same $\Gamma_k$'s,
but with different positive eigenvalues,
$\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of nonnegative numbers, $c_k$'s,
%
\begin{align}
&
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr)
\right)
%
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k \, \dot u_k \, \bigl( q(t) \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 2}^n \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k \, q(T) }
    \right]
\right)^2.
\label{eq:CSineq}
\end{align}
%
This inequality follows from the non-negativity of
the following quadratic polynomial of $Y$,
$$
\int_0^T
  dt \sum_{k = 2}^n \Gamma_k \,
    \left( \dot u_k \, Y - c_k \right)^2
  \equiv
  A \, Y^2 + B \, Y + C
%\ge 0,
$$
which implies a non-positive discriminant,
$B^2 - 4 \, A \, C$.
%
Note that the last expression of \eqref{eq:CSineq}
depends on the curve, $q(t)$ ($0 < t < T$),
only through the endpoint value, $q(T)$,
which is fixed in the variation process.
%
Thus, the inequality sets a lower bound
of the error $\Err$
given in Eq. \eqref{eq:error_asym1}.
%
The equality is achieved
if $\dot u_k\left( q(t) \right) = c_k$
for all $k = 2, \dots, n$ at any $t$,
up to a multiplicative factor.
%
Solving this set of equations yields
%
\begin{equation}
  \alpha(t) = \frac{              1             }
                   { \lambda_k \, (t + t_{k0} ) },
\label{eq:alpha_invtk}
\end{equation}
with
\begin{equation}
  t_{k0} = \frac{             T            }
                { e^{ \lambda_k q(T) } - 1 }.
\label{eq:tk0}
\end{equation}
%
Such a solution is possible only if
%
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
In this case,
all $t_{k0}$'s also share the same value,
$t_0$ by Eq. \eqref{eq:tk0},
such that
$u_k = (t + t_0) / (T + t_0)$,
and
$c_k \propto 1/(T + t_0)$
for $k \ge 2$.
%
The asymptotic error is then
\begin{align}
  E_A
  =
  \frac{       T     }
       { (T + t_0)^2 }
  \sum_{ k = 2 }^n
    \Gamma_k
  .
\label{eq:error_singlebin}
\end{align}
%
If we further assume that $\lambda_2 = 1$,
the optimal schedule given by Eq. \eqref{eq:alpha_invtk}
recovers Eq. \eqref{eq:alpha_invt1}.
%
\note{We add some steps for the above solution.
  Integrating $\dot u_k = c_k$ yields
  \begin{equation}
  u_k(t) = c_k \left(t + t_{k0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_k(T) = 1$ and $u_k(0) = e^{-\lambda_k \, q(T)}$,
  we have
  $$
  c_k = \frac{ 1 }{ T + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { T + t_{k0} }
  =
  e^{ -\lambda_k \, q(T) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_k \, q(T) + \lambda_k \, q(t)
  = \log c_k + \log\left( t + t_{k0} \right).
  $$
  Differentiating this with respect to $t$ yields
  $$
  \lambda_k \, \alpha(t)
  =
  \frac {     1      }
        { t + t_{k0} }.
  $$
}

Since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_k$,
we can freely set $\lambda_1$ to $\lambda_2$
without changing the nature of the updating scheme.
%
Now an updating matrix $\mathbf w$
with identical eigenvalues
is essentially a multiple of the identity matrix.
%
Thus, if no eigenvalue $\lambda_k$ is zero,
then in terms of asymptotic convergence,
the single-bin scheme adopted in the WL algorithm
is most efficient.
%

Note that
since we only optimized the asymptotic error,
the inverse-time schedule is optimal
only in the long time limit,
but not necessarily so for a finite-length simulation.
%
For example, consider a system that is initially
prepared under a fixed $\alpha_0$
at the beginning of the schedule $\alpha(t)$.
%
Then for shorter times, $T \approx 0$,
the error, dominated by the residual error
given by Eq. \eqref{eq:error_res1},
is clearly smaller if some eigenvalues
are less than $1.0$ (the single-bin case).
%
%In practice, real simulations
%may have long achieved the desired precision
%before reaching the asymptotic regime,
%especially if the updating matrix
%has some $\lambda_k$'s close to $0$.

\subsubsection{\label{sec:optscheme}
Cardinal updating schemes}

We can also generalize
the single-bin scheme to a class of
optimal updating schemes
that allow zero eigenvalues.
%
For example,
if we can assume that the PMF is sufficiently smooth
such that the error modes can be limited to $k \le K$.
%
Then we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_K = 1,
\mathrm{\; and \;}
\lambda_{K+1} = \cdots = \lambda_n = 0.
$$
%
Then,
Eq. \eqref{eq:error_singlebin} is modified to
$$
  E
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 2 }^K
    \Gamma_k.
$$
The optimal schedule of these schemes
remains to be the inverse-time schedule,
given by Eq. \eqref{eq:alpha_invt1}.
%
Below, we construct these schemes for
the homogeneous case.

For a periodic variable, if we demand
%
$$
\begin{aligned}
&
\lambda_1 = \cdots = \lambda_{K+1}
= \lambda_{n-K+1} = \cdots = \lambda_n = 1,
\\
&
\lambda_{K+2} = \cdots = \lambda_{n-K} = 0,
\end{aligned}
$$
for $K \ge 1$, then by using
Eq. \eqref{eq:mu_from_lambda_pbc},
we get
\begin{equation}
  \mu_l
  =
  \frac{
    \sin
    \dfrac{ (2 K + 1) \, l \, \pi }
         {              n        }
  }
  {
    n \, \sin \dfrac{ l \, \pi } { n }
  }
  .
  \label{eq:mu_cardinal_pbc}
\end{equation}
\note{Derivation.
$$
\begin{aligned}
\mu_l
&=
\frac 1 n \sum_{k = 0}^{n-1} \lambda_{k+1} \cos \frac{ k \, l \, 2 \, \pi } { n }
\\
&=
\frac{1}{n}
\left(
  1 +
  \sum_{k=1}^K
  \cos \frac { k \, l \, 2 \, \pi } { n }
  +
  \sum_{k=n-K}^{n-1}
  \cos \frac { k \, l \, 2 \, \pi } { n }
\right)
\\
&=
\frac 1 n
\sum_{k=-K}^K
\cos \frac { k \, l \, 2 \, \pi } { n }
=
  \frac{
    \sin
    \dfrac{ (2 K + 1) \, l \, \pi }
         {              n        }
  }
  {
    n \, \sin \dfrac{ l \, \pi } { n }
  }
.
\end{aligned}
$$
}

For a non-periodic variable, we demand
$$
\lambda_1 = \cdots = \lambda_{K+1} = 1,
\qquad
\lambda_{K+2} = \cdots = \lambda_n = 0.
$$
Then Eq. \eqref{eq:lambdan} gives
$\lambda_n = (-1)^{n+K-1}$,
and from Eq. \eqref{eq:mu_from_lambda_refl},
we have
\begin{equation}
  \mu_l
  =
  \frac{1}{2 \, n}
  \left[
    (-1)^{n+K+l-1}
    +
    \frac{
      \sin
      \dfrac{ (2 K + 1) \, l \, \pi }
           {         2 \, n        }
    }
    {
      n \, \sin \dfrac{ l \, \pi } { 2 \, n }
    }
  \right]
  .
\end{equation}
\note{Derivation.
$$
\begin{aligned}
  \lambda_{n+1}
  &=
  (-1)^{n-1}
  \left[
    \lambda_1
    + 2 (-\lambda_2 + \lambda_3 - \cdots + (-1)^K \lambda_{K+1})
  \right]
  \\
  &=
  \begin{dcases}
    (-1)^{n-1} & K \mathrm{\; even,} \\
    (-1)^n     & K \mathrm{\; odd.}
  \end{dcases}
\end{aligned}
$$
So
$$
\begin{aligned}
  \lambda_l
  &=
  \frac{1}{2\,n}
  \left[
    1 +
    2 \sum_{k=2}^{K+1}
    \cos \frac { (k - 1) \, l \pi } { n }
    +
    (-1)^{n+K-1} (-1)^l
  \right]
  \\
  &=
  \frac{1}{2\,n}
  \left[
    (-1)^{n+K+l-1}
    +
    \sum_{k=-K}^{K}
    \cos \frac { k \, l \pi } { n }
  \right]
  .
\end{aligned}
$$

Example 1. If $n = 3, K = 1$,
then $\lambda_1 = \lambda_2 = 1$, $\lambda_3 = 0$,
and $\lambda_4 = \lambda_1 - 2 \, \lambda_2 + 2 \, \lambda_3 = -1$.
%
We can verify that $\mu_0 = 1/3, \mu_1 = 1/2, \mu_2 = -1/6$.
%
Although $\mu_1 > \mu_0$, the system will not be trapped
in the current bin because of the reflective boundary condition.
%
Upon a visit to the middle bin, $j = 2$,
$v_i$ at bin $i = 1$ or $3$ is updated by $1/2 - 1/6 = 1/3$,
same as the update on bin $2$.
Upon a visit to bin $i = 1$, the three bins are updated
by $1/2 + 1/3 = 5/6$, $1/2 - 1/6 = 1/3$, and $-1/6$,
respectively.

Example 2. If $n = 3, K = 2$,
then $\lambda_1 = \lambda_2 = \lambda_3 = 1$, $\lambda_4 = 1$.
We have $\mu_0 = 1, \mu_1 = \mu_2 = 0$,
recovering the single-bin scheme.

Example 3. If $n = 4, K = 1$,
then $\lambda_1 = \lambda_2 = 1, \lambda_3 = \lambda_4 = 0$,
$\lambda_5 = (-1)(\lambda_1 - 2 \, \lambda_2 + 2 \, \lambda_3 - 2 \, \lambda_4 = 1$.
The kernel is given by
$\mu_0 = 1/2, \mu_1 = \sqrt{2}/8, \mu_2 = 1/4, \mu_3 = -\sqrt{2}/8$.
}

@@

To make better comparison with
the error of the single-bin case,
we shall rewrite the error as
a sum over the eigenmodes.
We define
\begin{align}
  \Theta_k
  =
  \frac{ 1 } { T }
  \sum_{ t = 1 } ^ T
    \theta_k( t )
  ,
\notag
\end{align}
%
with $\theta_k(t)$ defined by Eq. \eqref{eq:eta_def}
and $\phi_{k i}$ by Eq. \eqref{eq:eig_orthonormal_cols}.
%\begin{align}
%\theta_k( t )
%=
%\sum_{ i = 1 }^ n
%\phi_{k i} \left[
%             \frac{ h_i(t) }
%                  { p_i    }
%             - 1
%           \right],
%\end{align}
%and
%\begin{align}
%\sum_{ k = 1 }^n
%  \phi_{k i} \, \phi_{k j}
%  = p_i \, \delta_{i j}.
%\end{align}
%
Then,
%
\begin{align}
  E
  =
  \sum_{ k = 1 }^n
    \left\langle
      \Theta_k^2
    \right\rangle.
\notag
%\label{eq:error_sbin_equil}
\end{align}
%
\note{Because
%
\begin{align*}
\sum_{ k = 1 }^n
  \Theta_k^2
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ k = 1 }^n
    \theta_k( t ) \, \theta_k( t' )
\\
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ i, j = 1 }^n
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
    \left[
      \frac{ h_j(t') }
           { p_j     }
      - 1
    \right]
    \sum_{ k = 1 }^n
      \phi_{k \, i} \, \phi_{k \, j}
\\
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ i, j = 1 }^n
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
    \left[
      \frac{ h_j(t') }
           { p_j     }
      - 1
    \right]
    p_i \, \delta_{i, j}
\\
&=
\sum_{ i = 1 }^n
\frac{ 1 } { T }
\sum_{ t = 1 }^T
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
\frac{ 1 } { T }
\sum_{ t' = 1 }^T
    \left[
      \frac{ h_i(t') }
           { p_i     }
      - 1
    \right]
\\
&=
\sum_{ i = 1 }^n
  p_i \,
    \left(
      \frac{ \bar h_i }
           { p_i      }
      - 1
    \right)^2
.
\end{align*}
}
The sum can start from $k = 2$, for
by Eq. \eqref{eq:eigenmode1}, we have
$$
\theta_1(t)
=
\sum_{ i = 1 }^n
  \phi_{1i} \, \frac{ h_i(t) - p_i } { p_i }
=
\sum_{ i = 1 }^n h_i(t)
-
\sum_{ i = 1 }^n p_i = 0.
$$

@@

In conclusion,
we have proposed a method of computing
the optimal schedule of the updating magnitude
for a general adaptive FDS simulation.
%
Adaptive FDS methods
effectively sample a flat distribution
along a quantity of interest, $z$,
by constructing a bias potential along $z$
that offsets the PMF.
%
The optimal schedule ensures fastest convergence
of the bias potential to the PMF.
%They differ by the updating schemes,
%which specify how the bias potential is modified
%in each updating step.
%%
%In the single-bin or WL case,
%the update is limited to the current bin;
%while in the metadynamics case,
%the bias potential of a neighborhood of the current one
%is updated with relative weight specified by a Gaussian wave packet.
